Synthesizing Quotient Lenses

### 2018-03-05 papers/QuotientLenses/main.tex
Polishing and comments on early sections

% % Journal information % Supplied to authors by publisher for camera-ready
% submission; % use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{ICDP} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
While there currently exist many bidirectional programming languages, many of
them do not have built-in support for transformations that operate on data
modulo equivalence relations such as whitespace or permutation quotienting.
Moreover, langauges such as Boomerang which do allow for such transformations
are often difficult to program \sam{why?}. To address these shortcomings, this
paper (1) introduces a language of {\em quotient regular expressions} or QREs
for defining equivalences on regular expressions as well as a language of {\em
QRE lenses}, which are lenses that map between QREs, and (2) shows how to {\em
synthesize} QRE lenses from a pair of QREs and a set of example input-output
pairs. We also describe an implementation of QREs and QRE lens synthesis in
Boomerang, and demonstrate the utility of our implementation by using it to
synthesize QRE lenses between data formats drawn from {\tt data.gov} database.
\end{abstract}

@@

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate the same data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{ko2016bigul}, typically come with a variety of guarantees,
the \emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.
%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

One way to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
Returning to our bibliography example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.
%each of the data formats is quotiented by some equivalence relation and the lens
%laws are taken modulo equivalence.
%This allows lenses to be simpler and more
%flexible. For example, inessential details like whitespace and orderings of
%fields can be ignored when writing the transformations between the formats.

%However, while plain lenses are widely used in both standalone systems and
%popular libraries, Boomerang~\cite{boomerang} is the only implementation
%currently offering quotient lenses. Even though quotient lenses permit simpler
%transformations between the formats, they merely push complexity elsewhere.
Ordinary bijective lenses are fairly tricky to program, especially for
beginners, as they require programmers to think in two directions at once
and learn a new type system, which comes with some tricky ambiguity constraints.
Unfortunately, quotient lenses do not improve the situation:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we improve the process of programming quotient lenses
in two distinct, but connected ways. First, we
introduce a new language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives for strings that match the regular
expression. With
QREs, programmers do not need to define the data and canonization functions
separately---by adding simple annotations to ordinary regular expressions,
canonization functions can be inferred.

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses}, and, we show how to synthesize such QRE
lenses automatically.
More specifically, in
order to generate a lens, a programmer provides a pair of QREs
describing the source and target formats, and, optionally,
a suite of examples that demonstrates how the
lens should act on specific examples.  Given this information,
we are able to exploit and extend recent work on algorithms for synthesis of
bijective lenses~\cite{optician} to generate quotient lenses.
Importantly, the lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (a fact that cuts down the amount of non-determinism in
a specification and thus helps speed up synthesis), we prove that
this rigidity does not compromise expressiveness.

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for defining (1) a
  regular language modulo some equivalence relation, and (2) a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that our QREs are just as expressive as the full set of bijective
  quotient lenses.
  % In other words, if there eeach QRE lens $q$ defines the same transformation as a
  %QRE lens $q'$ that is in normal form.
  (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} discusses related work and~\ref{concl} summarizes the
results of the paper.

\section{Quotients by Example}
\label{sec:example}

@@

\emph{Lenses} define computations that can be read in two directions,
the get direction (forwards) and the put direction (backwards).  The simplest lenses define bijections between source and
data, but we are interested in \emph{quotient lenses}, which define
bijections modulo some equivalence relation.  The remains of this section
describe and motivate a new language and system for defining such lenses.

@@

When two instances of the same format differ only terms of such inessential
information, it is often useful to treat them as identical and to map them
into canonical representations.
Continuing our example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines},
}

@@

One natural way to define lenses that will synchronize data in these formats
is to provide the following items: \dpw{this is slightly redundant from the intro.  It
adds a little more info, but not much.  Should this be cut?  Should the intro be expanded
slightly?  Should it be left as is (more or less)?}

\begin{enumerate}
  \item A pair of descriptions of the external source and target formats.
  Such descriptions may be given by regular expressions, for instance.
  They serve as documentation to describe the domains over which the lens
  is to operate.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats, again, to document the structure of the transformation.
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.  In this case, there is
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and in between,
fields while also ignoring field order, one must nest canonizing combinators
within one another in a somewhat tricky way.

@@

We propose to do away with much of this overhead, and the burden of a
fiddly programming process, by defining a single new compact notation
to specify data formats modulo an equivalence relation and then by
synthesizing the lenses that map between these formats automatically.

@@

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$, an equivalence relation $\sim$ on the language
described by $R$, and a canonical set of representatives of $\sim$,
all at once.  QREs are constructed from a simple set of primitives
and composed using familiar regular operators
(choice `\cd{|}', concatenation `\cd{.}',
and Kleene star `$^*$').  Users may also treat QREs as pairs of
get and put functions and compose such pairs (the functional
composition operator `;').

For example, to begin construction of our \bibtex{} format, we might
first define a whitespace format, which, externally, matches
any non-zero number of whitespace characters, and converts
any such whitespace into a single space character (its canonical form).
Such a definition uses the \kw{project} primitive.

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in one format to the other.
For instance, assume
the \codefont{comma\_name} format describes ``Conway, J. H.''
and the \codefont{space\_name} format describes ``J. H. Conway''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

A third useful QRE primitive is the permutation combinator.  For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize any repeated separator between fields, one can extend a permutation
with any other QRE.  For instance, below, we normalize whitespace between fields,
leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
\end{lstlisting}
%
\noindent
The QRE for the bibtex format is presented in its entirety in Figure~\ref{fig:example-qre}.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}

%% In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
%% \verb+" "+ matches any non-zero number of whitespace characters and quotients
%% them to a single equivalence class with a single space as the canonical
%% representative. We call this QRE a \textit{projection} because an entire regular
%% language is projected to a single representative string.

%% Another QRE in Figure~\ref{fig:example-qre} is given by

%% \begin{center}
%% \begin{lstlisting}
%% squash comma_name $\to$ space_name using (get space_to_comma)
%% \end{lstlisting}
%% \end{center}

%% \noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
%% two data formats using a user-defined function. For example,
%% the QRE in the code above specifies that the \cd{comma_name} representation of
%% the author's name is equivalent to the \cd{space_name} representation, with
%% \cd{space_name} taken to be the canonical representation. The
%% function that identifies the two formats is given by \cd{get space_to_comma},
%% which transforms names in the \cd{comma_name} representation to names in the
%% \cd{space_name} representation. In this example, \cd{space_to_comma} is itself
%% an automatically \textit{synthesized} lens between the \cd{space_name} and
%% \cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
%% specify the squash transformation.

%% The third example of a QRE in the Figure is \cd{bib_fields}, defined by:
%% \begin{center}
%% \begin{lstlisting}
%% perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
%% \end{lstlisting}
%% \end{center}

%% \noindent This is a \textit{permutation} QRE that specifies that the label,
%% title, and author fields can occur in any order, with the fields separated by a
%% comma followed by arbitrary amounts of whitespace. The canonical representation
%% is that in which the label, author and title fields occur in that order,
%% separated by a comma followed by a newline. The \textit{permutation} QRE is
%% designed for equivalences such as this, where different permutations of
%% equivalences are interspersed with a separator which itself has an
%% equivalence defined on it.

%% Given the \textit{projection}, \textit{squash}, and \textit{permutation}
%% QRE primitives, we can build more complex languages using the usual regular
%% expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
%% and Kleene star `$^*$', and also sequential composition ``;''.

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get direction;
  produce the constant string $s$ in the put direction.
  \item \kw{ins} $s$: produce the string $s$ in the get direction;
  delete the string $s$ in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and by showing how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, a simpler subset of
lenses that support strict bijections, rather than bijections modulo
equivalence relations.

>>>>

% % Journal information % Supplied to authors by publisher for camera-ready
% submission; % use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{ICFP} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
While there currently exist many bidirectional programming languages, many of
them do not have built-in support for transformations that operate on data
modulo equivalence relations such as whitespace or permutation
quotienting.\bcp{I'm worried that most readers will not be able to follow
  that.  Moreover, this way of introducing the paper feels kind of
  negative---most languages are insufficient, and Boomerang is hard to
  program.  Could we say it in a more accessible, more positive way?}
Moreover, langauges such as Boomerang which do allow for such transformations
are often difficult to program \sam{why?}. To address these shortcomings, this
paper (1) introduces a language of {\em quotient regular expressions} or QREs
for defining equivalences on regular expressions as well as a language of {\em
QRE lenses}, which are lenses that map between QREs, and (2) shows how to {\em
synthesize} QRE lenses from a pair of QREs and a set of example input-output
pairs. We also describe an implementation of QREs and QRE lens synthesis in
Boomerang, and demonstrate the utility of our implementation by using it to
synthesize QRE lenses between a variety of real-world data formats used by
the {\tt data.gov} database. 
\end{abstract}

@@

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object.\bcp{These examples show that the same data {\em could} be
  represented in different formats, not that it needs to be.} When different applications manipulate the same data in different formats,
synchronizing the applications\bcp{Why are we talking about synchronizing
  applications?  Comes out of the blue.} amounts to synchronizing their
respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{ko2016bigul}\bcp{why mention exactly these two, out of all the
  bidirectional languages out there?}, typically come with a variety of
guarantees\bcp{grammatically, that says that each lens comes with a variety
  of guarantees---is that what you meant?},
the \emph{lens laws}, that help\bcp{why ``help''?} ensure put and get
functions are effective 
inverses\bcp{what is an effective inverse?} of one another.

One way\bcp{Do you mean to highlight the suggestion that there are other
  ways too?} to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
In bibliographies, for example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation\bcp{actually,
  two: one on each side; but more importantly ``introducing an equivalence
  relation'' is rather vague.} and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.

Ordinary bijective lenses are fairly tricky to program\bcp{do we have
  concrete evidence for this?}, especially for
beginners\bcp{or this?}, as they require programmers to think in two
directions at once\bcp{my own experience suggests that this is the less
  important issue...}
and learn a new type system, which comes with some tricky ambiguity
constraints\bcp{...and this is the bigger one.  But we don't explain what
  these ambiguity constraints are or why they must be imposed.}.
Unfortunately, quotient lenses do not improve the situation\bcp{Again, it
  feels rather negative.}:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we improve the process of programming quotient lenses
in two distinct, but connected ways. First, we
introduce a new language of {\em Quotient Regular Expressions} (QREs), which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives for strings that match the regular
expression. With
QREs, programmers do not need to define the data and canonization functions
separately---by adding simple annotations to ordinary regular expressions,
canonization functions can be inferred.

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses}, and, we show how to synthesize such QRE
lenses automatically.
More specifically, in
order to generate a lens, a programmer provides a pair of QREs
describing the source and target formats, plus 
a (possibly empty) suite of examples that demonstrates how the
lens should act in specific cases.  Given this information,
we can exploit and extend recent work on algorithms for synthesis of
bijective lenses~\cite{optician} to generate quotient lenses.
The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (which cuts down the amount of non-determinism in
a specification and greatly speeds up synthesis), we prove that
this rigidity does not compromise their expressiveness.

The following list summarizes the central contributions of our
work.\bcp{Ugh.} 
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that our QREs are just as expressive as the full set of bijective
  quotient lenses.
  (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\bcp{We've said that one of the main contributions of the paper is
  synthesizing lenses, but we don't actually show this happening in the
  introduction.  Should we?}

\section{QRE Lenses by Example}
\label{sec:example}

@@

\emph{Lenses} define computations that can be read in two directions, the
get direction (forwards) and the put direction (backwards).  The simplest
lenses define bijections between source and data, but we are interested in
\emph{quotient lenses}, which define bijections modulo some equivalence
relation\bcp{two!}.  The core ideas of quotient lenses were developed by
Foster et al~\cite{quotientlenses} and are embodied in the Boomerang
language.  This section describes and motivates a 
new notation for defining such lenses.

\bcp{We need two subsections here --- one on plain old bijective lenses, and
one on bijective lens synthesis.}

\subsection{Quotient Lenses}

@@

When two instances of the same format differ only terms of such inessential
information, it is useful\bcp{Would it be worth spelling out the alternative
(i.e., what lenses would look like if they had to preserve all this stuff?)}
to treat them as identical and to map them 
into canonical representations.
Continuing our example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines},
}

@@

One natural way to define lenses that will synchronize data in these formats
is to provide the following items: \dpw{this is slightly redundant from the intro.  It
adds a little more info, but not much.  Should this be cut?  Should the intro be expanded
slightly?  Should it be left as is (more or less)?} \saz{I think that this is
not too redundant and that it can be left as-is, modulo space crunching.}

\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
    \bibtex{} and EndNote.
  Such descriptions may be given by regular expressions, for instance.
  They serve as documentation to describe the domains over which the lens
  is to operate.\bcp{This way of saying it implies that they serve {\em
      only} as documentation.}
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats, again, to document the structure of the
  transformation.\bcp{Same question.}
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.\bcp{Is it really true
  that these types are optional in Boomerang?}  In this case, there is 
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.\bcp{Would it be worth
  demonstrating this concretely?}  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and between,
fields while also ignoring field order, one must nest canonizing combinators
in a somewhat tricky way.

@@

We propose to do away with much of this overhead, and the burden of a
fiddly programming process, by defining a single new compact notation
to specify data formats modulo an equivalence relation and then by
synthesizing the lenses that map between these formats automatically.
\bcp{If our main contribution is ``eliminating fiddliness,'' we should be
  more explicit about what the fiddliness is.  (But, again, I would be even
  happier if we could phrase everything more positively... ``Quotient lenses
  make lens programming more realistic.  But here are some opportunities for
  eliminating redundancy...'')}

@@

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$, an equivalence relation $\sim$ on the language described by
$R$, and a canonical set of representatives of $\sim$, all at once.  QREs
are constructed from a simple set of primitives and composed using familiar
regular operators (`\cd{|}' for choice, `\cd{.}' for concatenation, and
`$^*$' for repetition).  Users\bcp{This sentence feels clunky because it
  ``uses up'' its head/subject position, and needlessly focuses the reader's
  attention, on something (``users'') that is not especially important or
  germane to the conversation.  (There are lots of other sentences elsewhere
  in the paper that do the same thing. :-) Please rewrite (all of them).}
may also treat QREs as pairs of get and put functions and compose such pairs
(the functional composition operator `;').\saz{I agree with Benjamin about
  useless subjects, but here the sentence could be deleted entirely.}

For example, to begin construction of our \bibtex{} format, we might
first define a whitespace format, which, externally, matches
any non-zero number of whitespace characters, and converts
any such whitespace into a single space character (its canonical form).
This whitespace-normalizing QRE is easily defined by using the \kw{project} primitive: \bcp{Why is it called ``project''?}

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
  (throughout)?  That will make examples easier to read.}

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second. 
For instance, assume that
the \codefont{comma\_name} format describes ``Conway, J. H.''
and that the \codefont{space\_name} format describes ``J. H. Conway''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{s\_to\_c} canonization function used by squash
is simply to write it from scratch in some ordinary programming language. (Boomerang canonizers are written in OCaml.)
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{s\_to\_c} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
\bcp{Four problems here.  The first is that the notion of synthesizing lenses
comes out of nowhere (of course, it is in the background, but we're using it
here as if it had already been introduced concretely---this will be helped
by adding the new subsection I've suggested at the top of this section).
The second is that the idea of using half of a lens 
as a canonizer comes out of nowhere.  The third is that this idea is a hack (and
always was).  Why bother defining / synthesizing a lens and then using only
half of it?  And the fourth is that, in the original quotient lens paper,
we used this trick on lenses that were not even bijective, which may confuse
people familiar with the earlier paper.}The first line above synthesizes a
\emph{bidirectional}\bcp{Why is this italicized?  Are some lenses not
  bidirectional?} lens between 
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

Another QRE primitive is the sequential composition combinator ``;''.  For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \kw{squash} QRE and a QRE that projects away
the extra whitespace in between \codefont{first\_name} and \codefont{last\_name}

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}

\caption{QRE specification of \bibtex{} records. \bcp{The spacing of the
    comments is weird...}}
\label{fig:example-qre}
\end{figure}

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify that the
components of the permutation are conjoined by another QRE. For instance, below, we normalize whitespace
between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
\end{lstlisting}

% The final QRE combinator is the \kw{normalize} combinator.  This combinator is an
% ``escape hatch'' that allows a programmer to manually define a function $f$
% that sends each string matching a regular expression $R$ to some canonical
% representative in another regular expression $R'$. The equivalence relation
% defined by the normalize combinator is hence the equivalence relation defined by
% the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that match $R$,
% $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

% For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
% project QRE \kw{p\_wsp} which matches any non-zero number of whitespace
% characters and converts any such whitespace into a single space character could
% be expressed using the \kw{normalize} combinator as
% \begin{lstlisting}
% let p_wsp = normalize (wsp+, " ", f)
% \end{lstlisting}
% However, although it is in some sense ``universal'' (see the discussion below
% \saz{TODO}))the \kw{normalize} QRE imposes significant requirements on the
% canonization function $f$, so it is usually simpler to use the other QRE
% constructs where possible.

A QRE for the \bibtex{} format is presented in its
entirety in Figure~\ref{fig:example-qre}. \bcp{Walk through it a bit?}\saz{All of the
  examples in this section are walking through the figure, we just have to say so.}

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work\bcp{Why do we
  want to emphasize this here?}.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: \textit{delete} the constant string $s$ in the get direction;
  insert it in the put direction.
  \item \kw{ins} $s$: \textit{insert} the constant string $s$ in the get direction;
  delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
There is a fair amount of trickiness in writing these transformations, because
the programmer must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

\bcp{How about a subsection break here, to introduce the idea of QRE
  lens synthesis?}
